---
title: "A brief guide to GAReg"
author: "Mo Li, QiQi Lu, Robert Lund, Xueheng Shi"
date: "`r format(Sys.time(), '%d %b %Y')`"
output: 
  rmarkdown::html_vignette:
    number_sections: false
    toc: true
vignette: >
  %\VignetteIndexEntry{A brief guide to changepointGA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```


## Overview

**GAreg** implements genetic algorithm (GA) optimization for three core statistical workflows where discrete model structure must be learned from data:

1. Best subset variable selection – choose a parsimonious set of predictors under information or validation criteria.

2. Changepoint detection – estimate the number and locations of structural breaks (a.k.a. knots) in regression or time series.

3. Optimal spline knot selection – place interior knots for splines or piecewise polynomials to balance fit and smoothness.

The package provides a consistent GA interface, unified S4 results (gareg), and penalty-aware objectives (e.g., MDL, BIC/AIC, cross‑validation). It supports both varying‑knot and fixed‑knot modes, minimum spacing constraints, and unbalanced designs.

```{r setup}
library(GAReg)
```


## Best subset variable selection


## Changepoint detection


## Optimal spline knot selection

### Example data set

```{r}
library(MASS)

data(mcycle)
head(mcycle)

y = mcycle$accel
x = mcycle$times
x_unique = unique(x)
```

## Reproducibility

```{r}
sessionInfo()
```

## References
